Depression and suicidality as credible signals of need

### 2021-05-07 Paper.rmd
Clarified that verbal request was a signal per reviewer comments

This study utilized a between-subjects pretest-posttest design to examine how four different emotional signals (treatments) would influence (1) the degree participants believed a fictional character to be in need (*Belief*) and (2) the likelihood they would provide help (*Action*) relative to a simple *Verbal request* (the control condition), in four different vignettes, for a total of 20 conditions. In this design, the outcomes are measured at pre-treatment (T1). Participants are then randomized into either a control group or a treatment group, i.e., one of the emotional signals, and the outcomes are measured again (T2). Regression models (described later) are used to determine the effect of the treatment conditions on the posttest outcome variables, relative to the control condition, controlling for pretest levels of the outcome variables (we also explored within-subjects effects of the signal on outcomes at T2 compared to T1).

@@

Briefly, participants read a vignette that put them in situations in which an imagined sister was requesting a loan of \$50,000 from the participant that was originally saved for the participant's imagined daughter's college education. The sister claims the money is needed for medical treatment for her own child, but she has incentives to deceive. The manipulations were signals very similar to those used in this study, and the outcomes were belief that the sister really needed the money and intention to loan her the money. See XX for the full write-up. See <https://osf.io/3rg8b> for preregistration of the pilot study. See <http://doi.org/10.5281/zenodo.4637883> for pilot data.

@@

We learned six major lessons from the MTurk pilot study that influenced the design of the current study. First, as predicted, belief in the sister's need was highest in the *Depression* condition and lowest in the *Verbal request* condition, giving us confidence to continue this research (see Figure \@ref(fig:piloteffects)). Second, to rule out that believability and willingness to help were simply artifacts of the sister's psychiatric distress, we included a "signaling" condition in which the sister exhibited schizophrenic symptoms. As predicted, believability and willingness to help in this condition were dramatically lower than in any other condition, ruling out this alternative explanation. We therefore did not include the schizophrenic condition in the current study. Third, baseline (T1) believability was very high (mean = `r signif(pilot_results$meanT1belief, 2)` and median = `r signif(pilot_results$medianT1belief, 2)` on a 0-100 point scale), leaving little scope for signals to increase believability. In the current study we therefore endeavored to provide details in the vignettes that would undermine the victim's believability at baseline. Fourth, contrary to our hypothesis, the *Suicide threat* and *Suicide attempt* signals reduced believability relative to *Depression*. We surmised that the sister's adversity -- a sick child -- was not sufficient to justify a suicide attempt, thus undermining her believability. In the current study the signaler herself is therefore the victim, and the forms of adversity -- physical and sexual assault and thwarted and forced marriage -- are those that are frequently associated with suicidality. Fifth, the content of the vignette matters, so instead of a single vignette we decided to use four different vignettes to help determine the extent to which signaling effects were sensitive to seemingly irrelevant details of the vignettes. Sixth, in a randomized controlled trial, the control condition should be identical to the treatment condition in every respect except for the factor being tested, which in this study was signal cost. Post hoc, we realized that the control condition in the pilot study differed substantially from the treatment condition beyond signal cost: in the treatment conditions, participants encountered the sister twice: once when she requests help, and a second time when she signals. In our faulty control, participants only encountered the sister once, when she requests help, with no second encounter. In this study we therefore ensured that the control condition (Verbal request) was identical to the treatment conditions except for signal cost.

>>>>

This study utilized a between-subjects pretest-posttest design to examine how four different emotional signals (treatments) would influence (1) the degree participants believed a fictional character to be in need (*Belief*) and (2) the likelihood they would provide help (*Action*) relative to a simple *Verbal request* without additional signaling (the control condition), in four different vignettes, for a total of 20 conditions. In this design, the outcomes are measured at pre-treatment (T1). Participants are then randomized into either a control group or a treatment group, i.e., one of the emotional signals, and the outcomes are measured again (T2). Regression models (described later) are used to determine the effect of the treatment conditions on the posttest outcome variables, relative to the control condition, controlling for pretest levels of the outcome variables (we also explored within-subjects effects of the signal on outcomes at T2 compared to T1).

@@

Briefly, Participants read a vignette that put them in situations in which an imagined sister was requesting a loan of \$50,000 from the participant that was originally saved for the participant's imagined daughter's college education. The sister claims the money is needed for medical treatment for her own child, but she has incentives to deceive. The manipulations were signals very similar to those used in this study, and the outcomes were belief that the sister really needed the money and intention to loan her the money. See XX for the full write-up. See <https://osf.io/3rg8b> for preregistration of the pilot study. See <http://doi.org/10.5281/zenodo.4637883> for pilot data.

@@

We learned five major lessons from the MTurk pilot study that influenced the design of the current study. First, as predicted, belief in the sister's need was highest in the *Depression* condition and lowest in the *Verbal request* condition, giving us confidence to continue this research (see Figure \@ref(fig:piloteffects)). Second, to rule out that believability and willingness to help were simply artifacts of the sister's psychiatric distress, we included a "signaling" condition in which the sister exhibited schizophrenic symptoms. As predicted, believability and willingness to help in this condition were dramatically lower than in any other condition, ruling out this alternative explanation. We therefore did not include the schizophrenic condition in the current study. Third, baseline (T1) believability was very high (mean = `r signif(pilot_results$meanT1belief, 2)` and median = `r signif(pilot_results$medianT1belief, 2)` on a 0-100 point scale), leaving little scope for signals to increase believability. In the current study we therefore endeavored to provide details in the vignettes that would undermine the victim's believability at baseline. Fourth, contrary to our hypothesis, the *Suicide threat* and *Suicide attempt* signals reduced believability relative to *Depression*. We surmised that the sister's adversity -- a sick child -- was not sufficient to justify a suicide attempt, thus undermining her believability. In the current study the signaler herself is therefore the victim, and the forms of adversity -- physical and sexual assault and thwarted and forced marriage -- are those that are frequently associated with suicidality. Fifth, the content of the vignette matters, so instead of a single vignette we decided to use four different vignettes to help determine the extent to which signaling effects were sensitive to seemingly irrelevant details of the vignettes.
